---
title: "Exercises"
author: "Dieter"
date: "Last Updated: `r format(Sys.time(), '%d, %B, %Y at %H:%M')`"
output:
  github_document:
    html_preview: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# PART 1: regression, anova, and model comparison

For this series, we are using data from JSE Data Archive: <http://jse.amstat.org/jse_data_archive.htm>. I use the links to the data so that you can directly read the data from the internet (in most cases).

## Oysters

```{r}
data <-read_table('http://jse.amstat.org/datasets/30oysters.dat.txt')
colnames(data)
```

#### Predict weight by volume

-   Fit model
-   Plot results (scatter + regression line)

```{r}
result <- lm(Oyster_Weight_g ~ Oyster_Volume_cc, data = data)
summary(result)
```

```{r}
plot(data$Oyster_Volume_cc, data$Oyster_Weight_g)
abline(result, col = 'red')
```

## Heart rate

```{r}
data <- read_table('http://jse.amstat.org/datasets/normtemp.dat.txt', col_names = FALSE)
colnames(data) <- c('temp', 'gender', 'rate')
```

Variables:

-   Body temperature (degrees Fahrenheit)
-   Gender (1 = male, 2 = female)
-   Heart rate (beats per minute)

### Questions

-   Is there a significant difference between males and females in normal temperature?
-   Is there a relationship between body temperature and heart rate?
-   Compare your findings with a correlation analysis
-   Compare a model that only uses temperature to predict heart rate with one that also uses gender

Some additional questions:

-   Is the distribution of body temperatures normal?
-   Is the true population mean really 98.6 degrees F?
-   At what temperature should we consider someone's temperature to be "abnormal"?

### Examples


```{r}
model <- lm(temp ~ gender, data = data)
summary(model)
```


```{r}
data$gender <- as.factor(data$gender)
model <- lm(temp ~ gender, data = data)
summary(model)
boxplot(data$temp ~ data$gender)
```


```{r}
model <- lm(rate ~ temp, data = data)
summary(model)
```

```{r}
residuals <- model$residuals
hist(residuals)
```


```{r}
cor.test(data$rate, data$temp)
```


```{r}
plot(data$temp, data$rate)
abline(model)
```

```{r}
hist(data$temp)
```

## Home prices

-   NAME: Modeling home prices using realtor data
-   TYPE: Random sample
-   SIZE: 76 observations, 19 variables

DESCRIPTIVE ABSTRACT:

The data file contains information on 76 single-family homes in Eugene, Oregon during 2005. This dataset is suitable for a complete multiple linear regression analysis of home price data that covers many of the usual regression topics, including interaction and predictor transformations. Whereas realtors use experience and local knowledge to subjectively value a house based on its characteristics(size, amenities, location, etc.) and the prices of similar houses nearby, regression analysis can provide an alternative that more objectively models local house prices using these same data.

SOURCES:The data were provided by Victoria Whitman, a realtor in Eugene, in 2005. The data were used in a case study in Pardoe (2006).

VARIABLE DESCRIPTIONS:

-   id = ID number
-   Price = sale price (thousands of dollars)
-   Size = floor size (thousands of square feet)
-   Lot = lot size category (from 1 to 11)
-   Bath = number of bathrooms (with half-bathrooms counting as 0.1)
-   Bed = number of bedrooms (between 2 and 6)
-   BathBed = interaction of Bath times Bed
-   Year = year built
-   Age = age (standardized: (Year-1970)/10)
-   Agesq = Age squared
-   Garage = garage size (0, 1, 2, or 3 cars)
-   Status = act (active listing), pen (pending sale), or sld (sold)
-   Active = indicator for active listing (reference: pending or sold)
-   Elem = nearest elementary school (edgewood, edison, harris, adams,crest, or parker)

Edison = indicator for Edison Elementary (reference: EdgewoodElementary) Harris = indicator for Harris Elementary (reference: EdgewoodElementary) Adams = indicator for Adams Elementary (reference: EdgewoodElementary) Crest = indicator for Crest Elementary (reference: EdgewoodElementary) Parker = indicator for Parker Elementary (reference: EdgewoodElementary)

REFERENCES:Pardoe, I. (2006). Applied Regression Modeling: A Business Approach. Hoboken, NJ: Wiley.

```{r}
data <-read_table('http://jse.amstat.org/datasets/homes76.dat.txt')
```


### Questions

-   Does size or age influence price?
-   Do age and size interact in influencing the price?
-   Do houses near different school differ in price?

### Examples

```{r}
result <- lm(Y ~ X1, data = data)
summary(result)
```

```{r}
result <- lm(Y ~ elem, data = data)
summary(result)
```

## Cars

VARIABLE DESCRIPTIONS:

-   Price: suggested retail price of the used 2005 GM car in excellent condition. The condition of a car can greatly affect price. All cars in this data set were less than one year old when priced and considered to be in excellent condition.

-   Mileage: number of miles the car has been driven

-   Make: manufacturer of the car such as Saturn, Pontiac, and Chevrolet

-   Model: specific models for each car manufacturer such as Ion, Vibe, Cavalier

-   Trim (of car): specific type of car model such as SE Sedan 4D, Quad Coupe 2D

-   Type: body type such as sedan, coupe, etc.

-   Cylinder: number of cylinders in the engine\

-   Liter: a more specific measure of engine size

-   Doors: number of doors

-   Cruise: indicator variable representing whether the car has cruise control (1 = cruise)

-   Sound: indicator variable representing whether the car has upgraded speakers (1 = upgraded)

-   Leather: indicator variable representing whether the car has leather seats (1 = leather)

```{r}
library(readxl)
data <- read_xls('data/kuiper.xls')
colnames(data)
```

#### Questions

You know what to do:

1)  Assess the predictive value of one or more predictors on price.
2)  Construct two models and compare them

#### Examples

```{r}
model1 <- lm(Price ~ Mileage, data = data)
summary(model1)
```

```{r}
plot(data$Mileage, data$Price)
```

Ancova. We'll see this later...

```{r}
model2 <- lm(Price ~ Mileage + Make , data = data)
summary(model2)
```

Compare the models using the `anova function`.

```{r}
anova(model1, model2)
```
