---
title: "Simple Regression"
date: "Last Updated: `r format(Sys.time(), '%d, %B, %Y at %H:%M')`"
output:
  github_document:
    html_preview: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Functions we will use

+ `lm`, to fit a linear model
+ `summary`, to get the results from a linear model
+ `anova`, to get more results from a linear model
+ `predict`, to predict new values
+ `abline`, to plot a regression line


# Read Data

```{r}
library(tidyverse)
data <- read_csv('data/body.csv')
```

# Run simple regression model

```{r}
plot(data$Biiliac,data$Height)
```


```{r}
# Note: this syntax makes the predict function work
result <- lm(Height ~ Biiliac, data = data)
# This does not work with the predict function: result <- lm(data$Height ~ data$Biiliac)
summary(result)
```

If we wish to know how the F-value was calculate we can ask for the anova table.

```{r}
anova(result)
```
# Compare with correlation

```{r}
correlation <- cor.test(data$Height, data$Biiliac)
correlation
correlation$estimate^2
```

# Get the predicted values

```{r}
fitted <- fitted.values(result)
plot(data$Biiliac,data$Height)
points(data$Biiliac, fitted, col='red')
```

More general prediction:

```{r}

new_indepent_values <- c(10,15, 20, 25, 30, 35, 40)
new_data <- data.frame(Biiliac = new_indepent_values)

predicted <- predict(result, newdata=new_data)

plot(new_indepent_values,predicted, col='green', cex=2, pch=16)
points(data$Biiliac, fitted, col='red')
points(data$Biiliac, data$Height, col='black')

```
# Second example

```{r}
vik_data <- read_csv('data/vik_table_9_2.csv')
```
```{r}
plot(vik_data$X1, vik_data$Y)
result <- lm(Y ~ X1, data=vik_data)
summary(result)
```
# Easier regression line

```{r}
plot(vik_data$X1, vik_data$Y)
abline(result, col='red')
```

