---
title: "Exercises Prep"
author: "Dieter"
always_allow_html: true
output:
  github_document:
    html_preview: true
    toc: true
    toc_depth: 4
---

## Warning

**This document contains some of the solutions for the exercises. You're welcome to look over my shoulder but I encourage you to try the exercises for yourself first.**

```{r}
library(tidyverse)
```

## National Health and Nutrition Examination Survey

This is the [data source](https://wwwn.cdc.gov/nchs/nhanes/continuousnhanes/default.aspx?BeginYear=2017). See [here](https://wwwn.cdc.gov/Nchs/Nnyfs/Y_DEMO.htm) for more information about the demographic variables.

### Preprocess and merge

```{r}
library(haven)
demographics <- read_xpt('data/DEMO_J.XPT')
income <- read_xpt('data/INQ_J.XPT')
drugs <- read_xpt('data/DUQ_J.XPT')

drug_habits <- select(drugs, SEQN, DUQ200:DUQ280)
age<-select(demographics, SEQN, RIDAGEYR, RIDAGEMN)
finance <- select(income, SEQN, INDFMMPC,INDFMMPI, INQ300, IND235)
merged <- full_join(age, drug_habits, by='SEQN')
merged <- full_join(merged, finance, by='SEQN')
dim(merged)
```

Now, we need to do a bit of data cleaning. + The encoding for the cocaine use can be found [here](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/DUQ_J.htm#DUQ250). + The encoding for the income level can be found [here](https://wwwn.cdc.gov/Nchs/Nhanes/2017-2018/INQ_J.htm).

```{r}
merged <- mutate(merged, drug_used = case_when(DUQ200 == 1 ~ 1, TRUE ~ 0))
merged <- filter(merged, INDFMMPC<4)
```

Let's group the data

```{r}
grouped <- group_by(merged, INDFMMPC)
proportion <- summarise(grouped, count_coke = mean(drug_used, na.rm = TRUE))
plot(proportion$INDFMMPC, proportion$count_coke, type='b')
```
