---
title: "Assumptions"
date: "Last Updated: `r format(Sys.time(), '%d, %B, %Y at %H:%M')`"
output:
  github_document:
    html_preview: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Read some data

```{r}
library(tidyverse)
body_data <-read_csv('data/body.csv')
vik_data <-read_csv('data/vik_table_9_2.csv')
```

# Assumption 1: Linearity of model

## Example 1

```{r}
plot(vik_data$X1, vik_data$Y)
```

```{r}
model <- lm(Y ~ X1, data = vik_data)
plot(fitted(model), resid(model))
```
Let's create the diagnostic plots. For our purposes, we will only look at the top two plots.
```{r}
# Split the plotting panel into a 2 x 2 grid. Such that we get the four plots
# in 4 separate panels.
par(mfrow = c(2, 2))
plot(model)
# We can reset the plotting panel using this code: par(mfrow = c(1, 1))
```

## Example 2

```{r}
x_data <- runif(100)
y_data <- x_data^3 + rnorm(100, sd=0.05)
fake <- tibble(x_data=x_data, y_data = y_data)
model <- lm(y_data ~ x_data, data = fake)
par(mfrow = c(1, 1)) 
plot(x_data, y_data)
abline(model, col='red')
```

```{r}
model <- lm(y_data ~ x_data, data = fake)
summary(model)
```


```{r}
par(mfrow = c(2, 2))  # Split the plotting panel into a 2 x 2 grid
plot(model)
par(mfrow = c(1, 1)) 
```

Let's run the model on transformed data and look at the diagnostic plots again.

```{r}
model <- lm(y_data ~ I(x_data^3), data = fake)
par(mfrow = c(2, 2))  # Split the plotting panel into a 2 x 2 grid
plot(model)
par(mfrow = c(1, 1)) 
```


## Example 3

This demonstrates that the diagnostic plots can also be used when we have multiple predictors (in this case plotting the dependent vs the independents might be difficult or impossible).

```{r}
x1_data <- runif(100)
x2_data <- runif(100)
y_data <- x1_data^3 + x2_data^3
fake <- tibble(x1_data=x1_data,x2_data=x2_data, y_data = y_data)
model <- lm(y_data ~ x1_data + x2_data, data = fake)
par(mfrow = c(2, 2))  # Split the plotting panel into a 2 x 2 grid
plot(model)
par(mfrow = c(1, 1))
```

## Example 4

Finally, an example with real data.

```{r}
model <- lm(Weight ~ Height, data = body_data)
par(mfrow = c(2, 2))  # Split the plotting panel into a 2 x 2 grid
plot(model)
par(mfrow = c(1, 1)) 
```

# Assumption 2: Normal distribution of the errors

```{r}
model <- lm(Bicep ~ Shoulder, data = body_data)
residuals <- resid(model)
hist(residuals, 50)
```

```{r}
par(mfrow = c(2, 2))  # Split the plotting panel into a 2 x 2 grid
plot(model)
par(mfrow = c(1, 1)) 
```

# Assumption 3: Homoscedasticity

I created a data set with  heteroscedasticity. The gambling data below might be a good example as well.

```{r}
n<-500
errors <-rnorm(n, sd=seq(0.1,5, length.out=n))
x<-runif(n)
x<-sort(x)
y<- (20 * x) + errors
plot(x, y)
fake<-tibble(x=x, y=y)
model <- lm(y~x, data = fake) 

```

The model is still fitted properly!

```{r}
summary(model)
plot(x, y)
abline(model, col='red')
```

```{r}
par(mfrow = c(2, 2))  # Split the plotting panel into a 2 x 2 grid
plot(model)
par(mfrow = c(1, 1)) 
```

# More examples

## Heteroscedasticity: Gambling data

The teengamb data frame has 47 rows and 5 columns. A survey was conducted to study teenage gambling in Britain. This data frame contains the following columns:

+ sex 0=male, 1=female 
+ Socioeconomic status score based on parents’ occupation 
+ Income in pounds per week
+ Verbal score in words out of 12 correctly deﬁned
+ Gamble expenditure on gambling in pounds per year

```{r}
library(faraway)
data <- teengamb
head(data)
model<- lm(gamble~income, data=data)
```
```{r}
plot(data$income, data$gamble)
```
```{r}
summary(model)
```

```{r}
par(mfrow = c(2, 2))  # Split the plotting panel into a 2 x 2 grid
plot(model)
par(mfrow = c(1, 1)) 
```

```{r}
plot(data$status,data$gamble)
```

## Non-linear relationship: Pima data

The National Institute of Diabetes and Digestive and Kidney Diseases conducted a study on 768 adult female Pima Indians living near Phoenix. The dataset contains the following variables

+ Number of times pregnant
+ Plasma glucose concentration at 2 hours in an oral glucose tolerance test
+ Diastolic blood pressure (mm Hg)
+ Triceps skin fold thickness (mm) insulin 2-Hour serum insulin (mu U/ml)
+ Body mass index (weight in kg/(height in metres squared))
+ Diabetes pedigree function age Age (years)
+ Test whether the patient shows signs of diabetes (coded 0 if negative, 1 if positive)

```{r}
data<-pima
model <-lm(test ~ age + insulin, data=data)

par(mfrow = c(2, 2))  # Split the plotting panel into a 2 x 2 grid
plot(model)
par(mfrow = c(1, 1)) 
```

