---
title: "ggplot2"
author: "Dieter"
date: '2022-08-26'
output:
  github_document:
    html_preview: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Installing the gapminder package

We will use the `gapminder` package to get public health data. This data comes from the excellent [gapminder website](https://www.gapminder.org/). This package provides a `gapminder` tibble that contains the following variables:

+ country	
+ continent	
+ year	
+ lifeExp	life
+ pop
+ gdpPercap

Loading the package.

```{r}
library(tidyverse)
library(gapminder)
gap_data <- gapminder #So I can see the data in the environment tab
```

## A first simple example

```{r}
subset <- filter(gap_data, country == 'Algeria')
ggplot(subset) + aes(x=year, y=pop) + geom_point()
```
```{r}
subset <- filter(gap_data, country == 'Algeria')
ggplot(subset) + aes(x=year, y=pop) + geom_col() #Do not use geom_bar as that counts the data
```
We could use `geom_bar()` but we have to override the default counting behavior.

```{r}
subset <- filter(gap_data, country == 'Algeria')
ggplot(subset) + aes(x=year, y=pop) + geom_bar(stat= 'identity') 
```

```{r}
grp <- group_by(gap_data, country)
latest <- slice(grp, which.max(year))
grp <- group_by(latest, continent)
summary_data <- summarize(grp, sm = sum(pop) / 1000000000, mn = mean(pop))
p <- ggplot(summary_data) + aes(x = continent, y = sm) + geom_col()
p
ggsave('saved.pdf', p)
```
## Exploring some data

### Distributions

Read in some body data.

```{r}
data <- read_csv('data/body.csv')
```
Create a simple histogram

```{r}
ggplot(data) + aes(x = Hip, y=..density..) + geom_histogram() + geom_density(col='red')
```

Boxplots are also very nice tools to get a quick view of your data. Here, we overlay the raw data in red.

```{r}
ggplot(data) + aes(x = factor(Gender), y = Forearm) + geom_boxplot() + geom_jitter(width=0.1, height = 0, alpha = 0.25, col='red')
```
### Scatter plots

```{r}
ggplot(data) + aes(x = Hip, y = Forearm) + geom_point() + geom_smooth(method = "loess") +  geom_smooth(method = "lm", col='red')
```

## More aesthetics

Map variables on shapes and colors.

```{r}
#ggplot(data) + aes(x= Thigh, y=Waist, shape = Gender) + geom_point() # Does not work
#ggplot(data) + aes(x= Thigh, y=Waist, shape = as.factor(Gender), color= Gender) + geom_point() # Does work - but notice the output
ggplot(data) + aes(x= Thigh, y=Waist, shape = as.factor(Gender), color= as.factor(Gender)) + geom_point() # Does work - but notice the output
```
Another example using the gapminder data

```{r}
grp <- group_by(gap_data,year,continent)
mns <- summarise(grp, mn = sum(pop) / 1000000000)
ggplot(mns) + aes(x= year, y = mn, color = continent) + geom_point()
```
```{r}
subset <- filter(gap_data, country == 'Angola')
ggplot(subset) + aes(x= year, y = pop, size = gdpPercap) + geom_point()
```
## Small multiples: multiple panels

```{r}
grp <- group_by(gap_data, year, continent)
mns <- summarise(grp, mn = mean(lifeExp))
ggplot(mns) + aes(x = year, y=mn) + geom_point() + facet_grid(.~continent) 
```
```{r}
ggplot(mns) + aes(x = year, y=mn) + geom_point() + facet_wrap(~continent) + geom_smooth(method ='lm', color='gray', se=FALSE)
```

