---
title: "Simple Example of Linear Model"
date: "Last Updated: `r format(Sys.time(), '%d, %B, %Y at %H:%M')`"
output:
  github_document:
    html_preview: true
    toc: true
---

**This file lists the examples used in the slides.**

# A first example

Before starting download the following data sets:

+ `vik_table_9_2.csv`
+ `body.csv`

## Read and create data data

```{r}
data<-read.csv('data/vik_table_9_2.csv')

# Let's create two variables that could represent the age and weight of penguins
data$age<- 15 - data$X1
data$weight<-data$Y + 45
plot(data$age, data$weight)
```

## Run a simple linear model

```{r}
model1<-lm(weight ~ age, data=data) # fit this model: weight = B0 + B1 * age + error
summary(model1)
```

## Get the F value components

```{r}
anova(model1)
```

# Explicit model comparison

## Simple model

First, run the simple model...

```{r}
data$average_weight <- mean(data$weight) # Create a new variable which is the average of all weights
simple_model <- lm(weight ~ average_weight, data = data)
summary(simple_model)
```
## More complex model

Let's run the more complex model (again)...

```{r}
complex_model<-lm(weight ~ age, data=data)
```

## Compare the models

Compare the following output with that of the of the complex model.

```{r}
anova(simple_model, complex_model)
```

```{r}
anova(complex_model)
```


# Second example

```{r}
data <- read.csv('data/body.csv')
data<-data[1:20,]
head(data)
```

## Relationship between two variables: correlation

```{r}
correlation_test<-cor.test(data$Wrist, data$Age)
correlation_test
```

```{r}
correlation_test$estimate**2
```


## Relationship between two variables: linear model

Notice how the t-test for the parameter for `x` gives the same results as the t-test for the correlation test.

```{r}
model2<-lm(data$Wrist~data$Age)
summary(model2)
```


