---
title: "Solutions"
date: "Last Updated: `r format(Sys.time(), '%d, %B, %Y at %H:%M')`"
output:
  github_document:
    html_preview: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## One

Write a for loop that iterates over the numbers 1 to 7 and prints the cube of each number using print().

```{r}
for(i in 1:7){
  print(i^2)
  }
```

## Two

Write a while loop that prints out standard random normal numbers (use rnorm()) but stops (breaks) if you get a number bigger than 1.

Option 1

```{r}
value <- 0
counter <-0
while(value < 1)
{
  value <- rnorm(1)
  counter <- counter + 1
}
print(value)
print(counter)

```

Option 2
```{r}
counter <-0
while(TRUE)
{
  value <- rnorm(1)
  counter <- counter + 1
  if (value > 1){break}
}
print(value)
print(counter)

```

## Three

Using a for loop simulate the flip a coin twenty times, keeping track of the individual outcomes (1 = heads, 0 = tails) in a vector.

```{r}
repeats <- 20
outcomes <- character(repeats)
for(i in 1:repeats)
{
  outcome <- sample(c('H','T'), 1)
  outcomes[i] <- outcome
}
outcomes
```

You could do this in one line (but that was not the exercise).

```{r}
repeats <- 20
outcomes <- sample(c('H','T'), repeats, replace = TRUE)
outcomes
```

## Four

Use a while loop to investigate the number of terms required before the series 1 \times 2 \times 3 \times ,\ldots reaches above 10 million.

```{r}
product <- 1
term <- 0

while(product < 10000000)
{
  term <- term + 1
  product <- product * term
  
}
print(term)
# Check
1:term
cumprod(1:term)>10000000
```


 


