---
title: "Network Analysis"
author: "Dieter"
date: "Last Updated: `r format(Sys.time(), '%d, %B, %Y at %H:%M')`"
output:
  github_document:
    html_preview: true
    toc: true

---

## Loading the required packages

```{r}
library(sna)
library(igraph)
```

## Loading some data and create a gaph

```{r}
network_data <- read.delim('data/undirected.txt', header = FALSE, sep = " ")
graph1 <- graph_from_data_frame(network_data, directed = FALSE)
plot(graph1)
```

Also, notice the following functions:

+ `graph_from_adjacency_matrix()`
+ `graph_from_adj_list()`

## Create sample graphs

### Some simple functions

See: graph_from_literal(), make_chordal_ring(), make_empty_graph(), make_full_citation_graph(), make_full_graph(), make_graph(), make_lattice(), make_ring(), make_star(), make_tree()


```{r}
graph2 <- make_star(n=15, mode='in')
plot(graph2)
```

### Random graphs

These are often used to research the properties of graphs. See `sample_*` functions.

```{r}
sw <- sample_islands(3, 10, 5/10, 1)
plot(sw)
```

## Analyzing network properties


```{r}
sw <- sample_islands(3, 10, 5/10, 1)
degree(sw)
betweenness(sw)
edge_density(sw)
```



