---
title: "Basic Programming"
author: "Dieter"
output: 
  github_document:
    html_preview: true
    toc: true
---

## Basics of programming: variables and functions

Programming is basically (1) storing data, (2) performing operations on this data.

We will store data in so-called variables. We use functions to perform operations on the data.

## Basic operations: R as a calculator

R can perform the classic operations.

```{r}
1 / 200 * 30
(59 + 73 + 2) / 3
sin(pi / 2)
```

## Logical operations

```{r}
5 > 6
5 + 1 == 6 #NOTICE: I am using == to check equality!
1234 != 1234
```


## Storing data in variables

### Assigning data to a variable

R use `<-` to make assignments. This is a pain to type. You *could* use `=` but it will cause confusion later on.

```{r}
my_variable <- 5
```

Variables (also called values) come in many types (or classes). The very basic ones are the following:

```{r}
my_logical <- TRUE
my_character <- 'this is just a piece of text'
my_numeric <- 1.23455
```

These are very simple data types. We will often used much more complex ones when working with actual data.

```{r}
Name <- c("Jon", "Bill", "Maria", "Ben", "Tina")
Age <- c(23, 41, 32, 58, 26)
my_data_frame <- data.frame(Name, Age)
```

R studio shows `values` and `data` separately in the Environment window. However, this is just a visualization used by R studio. **You can use this window to inspect variables!**

### Note on naming variables

Try to use descriptive names for variables. And try to stick to a naming convention that works for you - preferably one that makes your code easy to read.

```{r}
i_like_snake_case <- 'snake_case'
otherPeopleUseCamelCase <- 'CamelCase'
some.people.use.periods <- 'periods.are.allowed'
And_aFew.People_RENOUNCEconvention <- 'Madness, Madness, I tell you!'
```

From R for Data Science:

*There\'s an implied contract between you and R: it will do the tedious
computation for you, but in return, you must be completely precise in
your instructions. Typos matter. Case matters.*

Also, it is important that you use names that are not keywords or functions in R. For example, the following is a bad idea:

```{r}
#length <- 15 ## THIS IS A BAD IDEA
```


### The vector

R is another basic variable in R. It's the simplest type of variable that actually allows you to store something recognizable as 'data'. We will spend some time on vectors as they are a good place to start to work with relatively simple data. Also, understanding how to work with vectors makes working with more complex data easier. Much of the operations you can do on vectors, which are 1D, can also be done on 2D data frames.

#### Creating a vector manually

```{r}
a_vector <- c(1, 5, 4, 9, 0)
another_one <- c(1, 5.4, TRUE, "hello")
```

#### Creating a vector using the `:` operator

```{r}
x <- 1:7
y <- 2:-2
```

#### Using `seq` to make a vector

```{r}
step_size <- seq(1, 10, by=0.25)
length_specified <- seq(1, 10, length.out = 20)
```

#### Indexing vectors

Every item in a vector has an index. Vector indices in R start from 1, unlike most programming languages where index start from 0.

```{r}
my_longer_vector <- c(1, 2, 'three', '4', 'V', 6, 7, 8)
```

You can use the `[]` to select (multiple) elements from a vector.

```{r}
my_single_element <- my_longer_vector[5]
the_start <- my_longer_vector[1:3]
my_part_of_vector <- my_longer_vector[c(1, 2, 5)] # I'm using a vector to select parts of a vector. Life is funny.
```

You can also use `[]` to overwrite a part of a vector

```{r}
my_longer_vector[1:3] <- c('replace', 'this', 'now')
```

#### Logical vectors

```{r}
vector1 <- c(1,5,6,7,2,3,5,4,6,8,1,9,0,1)
binary_vector <- vector1 > 5
binary_vector
```
```{r}
some_other_vector <- seq(from = 0, to = 100, length.out = length(vector1))
selected <- some_other_vector[binary_vector]
selected
```


## A trick

Before we go on, I want to share a simple trick. Using a IDE like Rstudio makes life easier (or at least it should). One of the benefits of the IDE is tab-completion. 

[DEMO GOES HERE]

## Functions

Now we know how to store data, we can start manipulating the data using functions. 

Functions take 0 or more inputs (also called arguments), perform some operation (i.e., the function of the function), and return some output. This output can be complex and consist of multiple parts. This are generic ways in which functions are used:

```
output <- function_name(arg1 = val1, arg2 = val2, ...)
output <- function_name(val1, val2, ...)
```

We've already encountered a function:

```{r}
output<-seq(from = 1, to= 123, by = 0.123)
```

How do we know which arguments a function can take? Using the help:

```{r}
?seq
```

Some very simple functions that might be usefull.

```{r}
a <- max(output)
b <- mean(output)
c <- min(output)
d <- ceiling(output)
e <- sd(output)
```








